# Reproducible Research: Assignment 1

##Loading and preprocessing the data

The data is read in. Here are the first ten rows of the data as well as the structure and a summary of the data:
```{r}
data<-read.csv("activity.csv")
head(data,10)
str(data)
summary(data)
```

The packages dplyr and ggplot2 will be needed.

```{r message=FALSE}
library(dplyr)
library(ggplot2)
```

As the column date is a factor variable, it is converted to a Date variable. Also the column interval indentifying the 5-minute interval in which the measurement was taken is an integer variable with for example 1655 meaning the time 4:55 pm. This is not suitable for further analysis and charts. Therefore another column is added showing the total minutes.

```{r}
data$date <- as.Date(data$date)
data<-mutate(data, minutes=floor(data$interval/100)*60+data$interval%%100)
```

Here is the structure and summary of the data after transforming it as decribed:
```{r}
str(data)
summary(data)
```


## What is the mean total number of steps taken per day?

### Total number of steps taken per day
```{r}
steps_per_day <- data %>% group_by(date) %>% summarize(total_steps=sum(steps))
steps_per_day
summary(steps_per_day$total_steps)
```

### Histogram of the total number of steps taken each day

``` {r}
hist(steps_per_day$total_steps,xlab="Total number of steps per day", 
     main="Total number of steps taken per day")
```

### The mean and median of the total number of steps taken per day

``` {r mean_median}
total_steps_mean <- mean(steps_per_day$total_steps,na.rm=TRUE)
total_steps_median <- median(steps_per_day$total_steps,na.rm=TRUE)
```
The mean of the total steps taken per day is `r total_steps_mean`, the median is 
`r total_steps_median`.

## What is the average daily activity pattern?

The following time series plot shows the average number of steps taken per 5-minute interval, averaged across all days.

```{r}
steps_per_interval <- data %>% group_by(minutes) %>% summarize(avg_steps_interval=mean(steps,na.rm=TRUE))
plot(steps_per_interval$minutes/60,steps_per_interval$avg_steps_interval,type='l',
     xlab='time in houres',ylab='number of steps',
     main='Average number of steps taken during one day')
```

```{r}
max_ind<-steps_per_interval$minutes[which.max(steps_per_interval$avg_steps_interval)]
max_ind_time <- paste0(floor(max_ind/60),":",max_ind%%60)
```
The maximum number of steps averaged across all days are taken in the interval `r max_ind` (in minutes) corresponding to time `r max_ind_time `.

## Imputing missing values

```{r}
steps_missing <- sum(is.na(data$steps))
```

There are `r steps_missing` missing values. 

These missing values are filled with the mean of the corresponding 5-minute interval. A new data set, data_new, is created with the missing data filled in.

```{r}
data2 <- merge(data,steps_per_interval,by.x="minutes",by.y="minutes")
missing <- is.na(data2$steps)
data2$steps[missing] <- data2$avg_steps_interval[missing]
data_new <- data2[,c("steps","date","interval","minutes")]
```
```{r}
steps_per_day_new <- data_new %>% group_by(date) %>% summarize(total_steps=sum(steps))
hist(steps_per_day_new$total_steps,xlab="Total number of steps per day", 
     main="Total number of steps taken per day with missing values filled in")

```
```{r}
total_steps_mean_new <- mean(steps_per_day_new$total_steps,na.rm=TRUE)
total_steps_median_new <- median(steps_per_day_new$total_steps,na.rm=TRUE)
```
The new mean of the total steps taken per day with the missing values filled in is `r total_steps_mean_new`, the median is `r total_steps_median_new`, which means that the mean and median are identical whereas they were slightly different when missing values were not filled in

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# note: Samstag is German for Saturday
data_new <- mutate(data_new,ind_weekday=ifelse(weekdays(data_new$date)>="Samstag","weekend","weekday"))
head(data_new)
steps_per_interval_weekday <- data_new %>% group_by(minutes,ind_weekday) %>% 
                                summarize(avg_steps=mean(steps)) 
g <- ggplot(steps_per_interval_weekday,aes(minutes/60,avg_steps)) + geom_line()+
            facet_wrap(~ind_weekday,nrow=2)+
            labs(title="Average number of steps taken during one day",
                 x="time in houres",y="number of steps")
print(g)
```